### 2018-03-29 Proposal.tex
Wrote project proposal

\documentclass[12pt]{article}
\usepackage{}
%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
\usepackage{times}
\usepackage{setspace}
\linespread{1.25}
\usepackage[letterpaper, margin=1in]{geometry}
\addtolength{\topmargin}{-0.5in}

\begin{document}
\title{Proposal: Predicting the Ending Notes of Bach Chorales}
\author{Nicholas Czaban}
\maketitle

In this project, I aim to create a model to predict the ending notes of Bach chorales. The model will be trained on data provided by University of California, Irvine's Machine Learning Repository. The data set contains the melody lines from 100 chorales, in a Lisp-formatted file. Every note of each chorale contains 6 fields of information: start-time of note, pitch of note, duration of note, key signature of note, time signature of note, and whether or not the note is under a fermata.\\

Bach chorales, as products of the Common Practice Period, follow a fairly well-defined set of rules in their compositions. However, there are enough variations allowable within the rules (as well as opportunities to subvert rules and expectations) to make the creation of a predictive model an interesting problem. Furthermore, the absence of bass, treble, or alto voices in the data set will require some extrapolation in order to create an accurate model. Given the ubiquity of Bach as an example of the period, such a model could provide useful information about the evolution of choral music over the years and track the eventual decline of the Common Practice style. In addition, this model would be able to characterize a composer's works, providing information about which keys and chord progressions they preferred.\\
\end{document}

### 2018-03-29 final.m
Fix cell within cell notation

@@ -14,10 +14,12 @@ clear tmp
 
 for j=1:length(result)
     for i=1:length(result{j})
-        if contains(result{j(1:i)}, '))((')
-           for k=length(result{j}):-1:i+1
-               result{j(1:k+1)} = result{j(1:k)};
-           end
+        pattern = '))((';
+        if strfind(result{j}{1:i}, pattern) > 0
+            disp "Found substring"
+%           for k=length(result{j}):-1:i+1
+%               result{j}(1:k+1) = result{j}(1:k);
+%           end
         end
 %         result{j(1:i)} = strrep(result{j(1:i)}, '(', '');
 %         result{j(1:i)} = strrep(result{j(1:i)}, ')', '');

### 2018-03-29 final.m
Fix cell within cell notation

@@ -12,10 +12,10 @@ for i=2:(length(result)/2)+1
 end
 clear tmp
 
+pattern = "st";
 for j=1:length(result)
     for i=1:length(result{j})
-        pattern = '))((';
-        if strfind(result{j}{1:i}, pattern) > 0
+        if contains(result{j}{1:i}, pattern) > 0
             disp "Found substring"
 %           for k=length(result{j}):-1:i+1
 %               result{j}(1:k+1) = result{j}(1:k);
@@ -25,3 +25,5 @@ for j=1:length(result)
 %         result{j(1:i)} = strrep(result{j(1:i)}, ')', '');
     end
 end
+
+clear i j

### 2018-03-29 final.m
Fix cell within cell notation

@@ -12,18 +12,19 @@ for i=2:(length(result)/2)+1
 end
 clear tmp
 
-pattern = "st";
+pattern = "))((";
 for j=1:length(result)
     for i=1:length(result{j})
-        if contains(result{j}{1:i}, pattern) > 0
-            disp "Found substring"
-%           for k=length(result{j}):-1:i+1
-%               result{j}(1:k+1) = result{j}(1:k);
-%           end
+        str=result{j}{i};
+        if strfind(str, pattern)>1
+            for k=length(result{j}):i+1:-1
+                result{j}{i+1} = result{j}{i};
+            end
+            
         end
-%         result{j(1:i)} = strrep(result{j(1:i)}, '(', '');
-%         result{j(1:i)} = strrep(result{j(1:i)}, ')', '');
+        result{j}{i} = strrep(result{j}{i}, '(', '');
+        result{j}{i} = strrep(result{j}{i}, ')', '');
     end
 end
 
-clear i j
+clear i j k pattern str

### 2018-04-03 AnnBib.tex
Wrote Annotated Bibliography

\documentclass[12pt]{article}
\usepackage{}
%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
\usepackage{times}
\usepackage{setspace}
\linespread{1.25}
\usepackage[letterpaper, margin=1in]{geometry}
\addtolength{\topmargin}{-0.5in}

\begin{document}
\title{Annotated Bibliography: Predicting the Ending Notes of Bach Chorales}
\author{Nicholas Czaban}
\maketitle

\noindent \hangindent=0.7cm Boyd, Malcolm. Bach, Oxford University Press, 2001. ProQuest Ebook Central, https://0-ebookcentral-proquest-com.dewey2.library.denison.edu/lib/wooster/detail.action?docID=281186.\\

This biography by Malcolm Boyd follows through Bach's life and looks at the various influences on the composer's music. Although the author has a very thorough section on the chorales Bach composed, his focus is more on the keyboard accompinament, rather than the vocalists themselves. Furthermore, there is not a great deal of in-depth analysis of the chorales in terms of their structure or progressions, although there are a few details mentioned here and there. However, the broader focus of the book will be able to back up or refute any inferences to be gained from my analysis of the chorale data.\\

\noindent \hangindent=0.7cm De Felice, Clelia, et al. "Splicing music composition." Information Sciences, vol. 385 - 386, 2017, pp. 196-212. OhioLINK Electronic Journal Center, doi:10.1016/J.INS.2017.01.004.\\

In this article, the authors suggest a unique application of splicing systems, a model originally concieved to study DNA recombination. The authors apply the methodology of this splicing system to train a computer to compose new music. The authors also calculated the overall probability of a particular chord progression being used. This source will be an excellent reference for cross-verification. Additionally, the computer-generated pieces could be used in an extension of the project once the model is complete.\\

\noindent \hangindent=0.7cm Dua, D. and Karra Taniskidou, E. (2017). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.\\

This data set contains the melody lines from 100 Bach chorales, containing information about each note in the piece. Every note of each chorale contains 6 fields of information: start-time of note, pitch of note, duration of note, key signature of note, time signature of note, and whether or not the note is under a fermata.\\

\noindent \hangindent=0.7cm Miranda, Eduardo R. Composing music with Computers. New York, NY: Routledge, 2014. Safari Books Online, http://0-proquest.safaribooksonline.com.dewey2.library.denison.edu\\/book/audio/9780240515670.\\

This comprehensive text on autonomous composition spends much of its time on the necessary math and computer science knowledge needed to create an automated composer. It covers two major techniques: nueral computation and iterative algorithms. Both of these methods could provide excellent starting points for developing my own model.\\

\noindent \hangindent=0.7cm Ultan, Lloyd. Music Theory : Problems and Practices in the Middle Ages and Renaissance, University of Minnesota Press, 1973. ProQuest Ebook Central, https://0-ebookcentral-proquest-com.dewey2.library.denison.edu/lib/wooster/detail.action?docID=316566.\\

This text provides excellent research on the development of music during the Middle Ages and Renaissance, leading up to and including the early portion of the common practice period. Each chapter focuses on the chronological development of various conventions, with sections on both the musical devices and the larger format of pieces.
\end{document}

### 2018-04-03 final.m
Split lisp file into Matlab-readable data

@@ -12,19 +12,57 @@ for i=2:(length(result)/2)+1
 end
 clear tmp
 
-pattern = "))((";
+data(100, 999) = "NaN";
+pattern = '))((';
 for j=1:length(result)
+    k=1;
     for i=1:length(result{j})
-        str=result{j}{i};
+        str = result{j}{i};
         if strfind(str, pattern)>1
-            for k=length(result{j}):i+1:-1
-                result{j}{i+1} = result{j}{i};
-            end
-            
+            tmp = regexp(str, '))((', 'split');
+            data(j,k) = tmp{1};
+            k=k+1;
+            data(j,k) = tmp{2};
+        else 
+            data(j,k) = str;
         end
-        result{j}{i} = strrep(result{j}{i}, '(', '');
-        result{j}{i} = strrep(result{j}{i}, ')', '');
+        k=k+1;
     end
 end
 
-clear i j k pattern str
+for j=1:100
+    for i=1:1033
+        data(j,i) = strrep(data(j,i), '(', '');
+        data(j,i) = strrep(data(j,i), ')', '');
+    end
+end
+
+clear i j k pattern str result tmp
+
+%% Separate data by chorale
+for j=1:100
+    i=0;
+    k=1;
+    n=true;
+    while n==true
+        k=k+1;
+        if k>length(data)
+            n=false;
+        elseif data(j,k) == "st"
+            i=i+1;
+            k=k+1;
+            chorale(i,1) = data(j,k);
+            k=k+2;
+            chorale(i,2) = data(j,k);
+            k=k+2;
+            chorale(i,3) = data(j,k);
+            k=k+2;
+            chorale(i,4) = data(j,k);
+            k=k+2;
+            chorale(i,5) = data(j,k);
+            k=k+2;
+            chorale(i,6) = data(j,k);
+        end
+    end
+end
+clear n ans i j k

### 2018-04-05 final.m
Fix parsing error

@@ -43,12 +43,8 @@ clear i j k pattern str result tmp
 for j=1:100
     i=0;
     k=1;
-    n=true;
-    while n==true
-        k=k+1;
-        if k>length(data)
-            n=false;
-        elseif data(j,k) == "st"
+    while k<=length(data)
+        if data(j,k) == "st"
             i=i+1;
             k=k+1;
             chorale(i,1) = data(j,k);
@@ -63,6 +59,34 @@ for j=1:100
             k=k+2;
             chorale(i,6) = data(j,k);
         end
+        k=k+1;
     end
+    % Create Naive Model - Assumes that every chorale ends on the tonic of
+    % its key
+    chorale=str2double(chorale);
+    N_train = chorale(1:size(chorale,1)-1, :);
+    N_val = chorale(size(chorale,1), :);
+    switch(N_val(4))
+        case 0 % C Major
+            prediction=60;
+        case 1 % G Major
+            prediction=67;
+        case 2 % D Major
+            prediction=62;
+        case 3 % A Major
+            prediction=69;
+        case 4 % E Major
+            prediction=64;
+        case -1 % F Major
+            prediction=65;
+        case -2 % Bb Major
+            prediction=70;
+        case -3 % Eb Major
+            prediction=63;
+        case -4 % Ab Major
+            prediction=68;
+    end
+    
+    clear chorale
 end
-clear n ans i j k
+clear n ans i j k data

### 2018-04-17 Paper.tex
Start paper

\documentclass[12pt]{article}
\usepackage{}
%%%%%%%%%%%%%%%%%%%%%%%%
% PREAMBLE
\usepackage{times}
\usepackage{setspace}
\linespread{1.25}
\usepackage[letterpaper, margin=1in]{geometry}
\addtolength{\topmargin}{-0.5in}

\begin{document}
\title{Predicting the Ending Notes of Bach Chorales}
\author{Nicholas Czaban}
\maketitle

\section{Introduction}
It is sometimes said that the key difference between computers and humans is a human's artistry, in both creation and enjoyment of the arts. Many studies have attempted to challenge this view, with the development of artificial intelligence that can compose original music. However, this paper will focus on the other perspective of music, that of the listener. With exposure to the vast canon of musical history, humans form their own expectations about the ``rules'' of music; even if someone knows nothing of music theory, they could internally recognize an unresolved note or a dissonant interval. Similarly, this paper will describe a computer model which predicts the ending note of various melodies taken from a set of Bach chorales.
\section{Computer-Generated Music}
There have been a number of attempts to create algorithmic composition software, a program which can compose music with little human interaction or involvement; one of the first examples dating to the 1950's when Lejaren Hiller and Leonard Isaacson produced {\it The Illiac Suite for String Quartet} \cite{computers}. There are two varieties of algorithmic composition: top-down and bottom-up; an implementation of the bottom-up approach is covered in detail later.\\

A bottom-up approach tasks the computer with composing a number of distinct passages \cite{computers}. A human then examines the output and stores any phrases that they deem promising. The sections are later combined into a single work, with or without additional input from the computer. The top-down approach involves a composer detailing a computer program which mirrors the composer's specific plan for the composition. Top-down approaches lend themselves less to algorithmic composition, but there are exceptions; most top-down implementations are used as tools in the same sense as a word processor.
\subsection{A Bottom-Up Approach: Splicing Systems}
In their attempt to create an bottom-up algorithmic composition model, De Felice et. al. use a technique originally concieved for DNA analysis called a {\it splicing system} to teach a computer to compose original music in the style of Bach \cite{computer_composer}. After converting the chorales into their own tonality degree representation (a notation system that records the voice, pitch, octave, and major/minor tonality of each note), each word is passed into the splicing system. It is in this splicing system that the harmonic function (i.e. the chord progression) and the melodic function (how each note relates to the chord) are developed from the input information. The final model uses a combination of predefined rules of classical music and statistical information extracted from existing data in its compositions.
\section{Data}
The data set used to test the models is a collection of one hundred Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}. Each chorale contains only the melody line from each chorale. Each note in the melody has six different information fields. The starting time of the note is measured in 16th notes from the start of the piece. The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C\#4 given a value of 61, etc. The pitches are restricted to values between 60 and 75, by nature of the notes present in the chorales. The duration of each note is given, also measured in 16th notes. The key signature is presented as values between -4 and +4. The numbering relates to the number of sharps or flats are present in the key signature: positive for sharps, negative for flats. The time signature is again measured in the number of 16th notes, this time in the number of 16th notes per measure. Finally, a boolean flag is set to true if the note is underneath a fermata.
\section{Methods}
Three separate models were created for this experiment: a naive baseline model, a human-made model with predefined rules, and a computer-generated model.
\subsection{Baseline Model}
The baseline model relied on one simple rule: the final note of the melody would be the root of the key in which the chorale began. This model was applied to all the melodies in the data set. Overall, this approach was able to correctly predict the note in 49 of 100 chorales. Of the incorrect predictions, 36 chorales ended on a note below the predicted note and 15 ended on a note above the predicted note.\\

Despite being a baseline model, the near 50\% success rate indicates the relative simplicity of the problem at hand. This model works in a few select cases: either the chorale does not change keys, or the root is included in the new key. Given that key signatures do not contain every note of the octave, the first case is more likely to occur than the second. Thus, these initial results would suggest that there is a common pattern among Bach's choral pieces; that a plurality of his work did not deviate in key.
\subsection{Human Model}
To predict the final note, the smart model used the two preceeding notes as reference. Looking at the key signature and the pitch, the model performs modulo operations on the MIDI pitch to find the relative position in the key, with values between 0 and 11. For example, the tonic becomes 0, the major third becomes 4, and the perfect fifth becomes 7. With these values calculated, the model uses a defined set of patterns to predict the melody endings. For the sake of simplicity, only notes in the major triad were used in this model's predictions.\\

This model is intended to mimic how a passive listener might try to predict the ending. The patterns were not directly based on the rules of music theory or the Common Practice period, but instead relied on a combination of what would sound least dissonant to the listener: the nearest note in the triad. For instance, any chorale where the penultimate note was the major seventh predicted the last note to be the root of the key. The same note would be predicted if the model was passed the major third and major second of the key.\\

This model predicted the endings of a total of 56 chorales correctly. While the number of overestimations was reduced by around 60\%, there was a 50\% rise in the number of underestimations. The minimal inprovement of the model would suggest that the patterns created for this model were either incorrect or insufficiently described the musical patterns. In any case, there is much room for improvement for this model.
\subsection{Computer Model}
In the final model, each note of the chorale was analyzed separately, then given a probability of that pitch being the same note as the final note. The 70 random chorales from the data set were used as the training set, with the remaining 30 chorales left for testing.
\section{Conclusion}

\newpage
\begin{thebibliography}{10}
\bibitem{computer_composer}
  De Felice, Clelia, et al. "Splicing music composition." Information Sciences, vol. 385 - 386, 2017, pp. 196-212. OhioLINK Electronic Journal Center, doi:10.1016/J.INS.2017.01.004.

\bibitem{data_set}
  Dua, D. and Karra Taniskidou, E. (2017). UCI Machine Learning Repository [http://archive.ics.uci.edu/ml]. Irvine, CA: University of California, School of Information and Computer Science.

\bibitem{computers}
  Miranda, Eduardo R. Composing music with Computers. New York, NY: Routledge, 2014. Safari Books Online, http://0-proquest.safaribooksonline.com.dewey2.library.denison.edu\\/book/audio/9780240515670.
\end{thebibliography}
\end{document}

### 2018-04-17 Paper.tex
Finished first draft

@@
 With exposure to the vast canon of musical history, humans form their own expectations about the ``rules'' of music; even if someone knows nothing of music theory, they could internally recognize an unresolved note or a dissonant interval.
-Similarly, this paper will describe a computer model which predicts the ending note of various melodies taken from a set of Bach chorales.
+Similarly, this paper will describe a computer model which predicts the ending note of various melodies taken from a set of Bach chorales.\\
+
+Johann Sebastian Bach, as a product of the Baroque period, followed a set of composition guidelines which gave the Common Practice Period its name.
+These guidelines dealt with methods of resolving dissonance, writing melodies, and more.
+Despite decades and centuries of experimentation since Bach's death in 1750, many of these common practices can still be seen in Western music.
+Thus, his work makes for an excellent starting point for predictive modeling.
+
 \section{Computer-Generated Music}
@@
 Top-down approaches lend themselves less to algorithmic composition, but there are exceptions; most top-down implementations are used as tools in the same sense as a word processor.
+
 \subsection{A Bottom-Up Approach: Splicing Systems}
@@
 al.
-use a technique originally concieved for DNA analysis called a {\it splicing system} to teach a computer to compose original music in the style of Bach \cite{computer_composer}.
+use a technique originally conceived for DNA analysis called a {\it splicing system} to teach a computer to compose original music in the style of Bach \cite{computer_composer}.
 After converting the chorales into their own tonality degree representation (a notation system that records the voice, pitch, octave, and major/minor tonality of each note), each word is passed into the splicing system.
@@
 the chord progression) and the melodic function (how each note relates to the chord) are developed from the input information.
-The final model uses a combination of predefined rules of classical music and statistical information extracted from existing data in its compositions.
+The final model uses a combination of predefined rules of classical music and statistical information extracted from existing data in its compositions.\\
+
+The authors determined weights for each possible chord progression based on their frequency in common practice \cite{computer_composer}.
+These weights incorporate both progressions within a single key as well as progressions that change keys.
+
 \section{Data}
@@
 The time signature is again measured in the number of 16th notes, this time in the number of 16th notes per measure.
-Finally, a boolean flag is set to true if the note is underneath a fermata.
+Finally, a Boolean flag is set to true if the note is underneath a fermata.
+
 \section{Methods}
 Three separate models were created for this experiment: a naive baseline model, a human-made model with predefined rules, and a computer-generated model.
+
 \subsection{Baseline Model}
@@
 Thus, these initial results would suggest that there is a common pattern among Bach's choral pieces; that a plurality of his work did not deviate in key.
+
 \subsection{Human Model}
-To predict the final note, the smart model used the two preceeding notes as reference.
+To predict the final note, the smart model used the two preceding notes as reference.
 Looking at the key signature and the pitch, the model performs modulo operations on the MIDI pitch to find the relative position in the key, with values between 0 and 11.
@@
 This model predicted the endings of a total of 56 chorales correctly.
-While the number of overestimations was reduced by around 60\%, there was a 50\% rise in the number of underestimations.
-The minimal inprovement of the model would suggest that the patterns created for this model were either incorrect or insufficiently described the musical patterns.
+While the number of over-estimations was reduced by around 60\%, there was a 50\% rise in the number of underestimations.
+The minimal improvement of the model would suggest that the patterns created for this model were either incorrect or insufficiently described the musical patterns.
 In any case, there is much room for improvement for this model.
+
 \subsection{Computer Model}
@@
 The 70 random chorales from the data set were used as the training set, with the remaining 30 chorales left for testing.
+
 \section{Conclusion}
+Regardless of the results of the model, the relationship between computers and music continues to provide a challenging roadblock on the path to human-like artificial intelligence.
+The amount of interdisciplinary cooperation that will eventually be necessary is staggering; from computer scientists and mathematicians to create the models, to musicians for the actual art, to psychologists and philosophers to determine how a computer can best ``learn.''
 
-\newpage
 \begin{thebibliography}{10}

### 2018-04-17 Paper.tex
Completed computer model

@@
 This model predicted the endings of a total of 56 chorales correctly.
-While the number of over-estimations was reduced by around 60\%, there was a 50\% rise in the number of underestimations.
+While the number of overestimations was reduced by around 60\%, there was a 50\% rise in the number of underestimations.
 The minimal improvement of the model would suggest that the patterns created for this model were either incorrect or insufficiently described the musical patterns.
@@
 \subsection{Computer Model}
-In the final model, each note of the chorale was analyzed separately, then given a probability of that pitch being the same note as the final note.
-The 70 random chorales from the data set were used as the training set, with the remaining 30 chorales left for testing.
+In the final model, each chorale was used in a decision tree to predict the final pitch.
+70 random chorales from the data set were used as the training set, with the remaining 30 chorales left for testing.
+This model was able to correctly predict between 62 and 75 of the 100 chorales.
 

### 2018-04-17 final.m

@@ -186,10 +186,70 @@ for j=1:100
         case -4 % Ab Major
             results(j,3)=mod(val(1,2)-8, 12);
     end
+    
+    validation(j,:)=val(2);
     clear chorale val train n1 n2
 end
 z=zeros(100,1);
-confusionmat(z, results(:,1))
-confusionmat(z, results(:,2))
-confusionmat(results(:,3), tmp(:,2))
-clear z n ans i j k data
+%confusionmat(z, results(:,1))
+%confusionmat(z, results(:,2))
+    % Confusion Matrix based on the relative position in the key
+%confusionmat(results(:,3), tmp(:,2))
+clear z n ans i j k
+
+%% Build computer model: For each note in the chorale, calculate the probability of that note being the ending pitch
+for j=1:100
+    k=2;
+    i=1;
+    while k<=length(data)
+        if data(j,k) == "st"
+            k=k+1;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            chorale(j,i) = data(j,k);
+            i=i+1;
+        elseif j==1
+            chorale=str2double(chorale);
+            break;
+        else
+            break;
+        end
+        k=k+1;
+    end
+end
+clear i j k
+k=randperm(100);
+train=chorale(k(1:70),:);
+val=validation(k(1:70),:);
+T=fitctree(train, val);
+P=predict(T, chorale);
+confusionmat(P, validation)
+clear k
+j=0;
+over=0;
+under=0;
+for i=1:100
+    if P(i)-validation(i)==0
+        j=j+1;
+    elseif P(i)-validation(i)>0
+        over=over+1;
+    else
+        under=under+1;
+    end
+end
+view(T,'Mode', 'Graph')
+computer_model = [over;j;under]
+clear i

### 2018-04-17 final.m
Adding tree diagram to paper

@@ -251,5 +251,5 @@ for i=1:100
     end
 end
 view(T,'Mode', 'Graph')
-computer_model = [over;j;under]
-clear i
+computer_model=[over;j;under]
+clear i j under over P data ans train val computer_model

### 2018-04-23 final.m
Fixed computer model, redrew tree

@@ -231,20 +231,45 @@ for j=1:100
     end
 end
 clear i j k
-k=randperm(100);
-train=chorale(k(1:70),:);
-val=validation(k(1:70),:);
+train=chorale(21:100, :);
+val=validation(21:100, :);
 T=fitctree(train, val);
-P=predict(T, chorale);
-confusionmat(P, validation)
-clear k
+P=predict(T, chorale(1:20, :));
+finalResults=P;
+
+train=chorale([1:20 41:100], :);
+val=validation([1:20 41:100], :);
+T=fitctree(train, val);
+P=predict(T, chorale(21:40, :));
+finalResults(21:40, :)=P;
+
+train=chorale([1:40 61:100], :);
+val=validation([1:40 61:100], :);
+T=fitctree(train, val);
+P=predict(T, chorale(41:60, :));
+finalResults(41:60, :)=P;
+
+train=chorale([1:60 81:100], :);
+val=validation([1:60 81:100], :);
+T=fitctree(train, val);
+P=predict(T, chorale(61:80, :));
+finalResults(61:80, :)=P;
+
+train=chorale(1:80, :);
+val=validation(1:80, :);
+T=fitctree(train, val);
+P=predict(T, chorale(81:100, :));
+finalResults(81:100, :)=P;
+
+confusionmat(finalResults, validation)
+clear k val train
 j=0;
 over=0;
 under=0;
 for i=1:100
-    if P(i)-validation(i)==0
+    if finalResults(i)-validation(i)==0
         j=j+1;
-    elseif P(i)-validation(i)>0
+    elseif finalResults(i)-validation(i)>0
         over=over+1;
     else
         under=under+1;

### 2018-04-23 final.m
Create presentation

@@ -190,12 +190,10 @@ for j=1:100
     validation(j,:)=val(2);
     clear chorale val train n1 n2
 end
-z=zeros(100,1);
-%confusionmat(z, results(:,1))
-%confusionmat(z, results(:,2))
-    % Confusion Matrix based on the relative position in the key
-%confusionmat(results(:,3), tmp(:,2))
-clear z n ans i j k
+
+% confusionmat(N_results(:,1), N_results(:,2))
+confusionmat(tmp(:,1), N_results(:,2))
+clear n ans i j k
 
 %% Build computer model: For each note in the chorale, calculate the probability of that note being the ending pitch
 for j=1:100

### 2018-04-24 final.m
Tried K nearest neighbors, which did not work well

@@ -190,9 +190,11 @@ for j=1:100
     validation(j,:)=val(2);
     clear chorale val train n1 n2
 end
-
+z=zeros(1,100);
+confusionmat(z, results(:,1))
+confusionmat(z, results(:,2))
 % confusionmat(N_results(:,1), N_results(:,2))
-confusionmat(tmp(:,1), N_results(:,2))
+% confusionmat(tmp(:,1), N_results(:,2))
 clear n ans i j k
 
 %% Build computer model: For each note in the chorale, calculate the probability of that note being the ending pitch
@@ -233,46 +235,46 @@ train=chorale(21:100, :);
 val=validation(21:100, :);
 T=fitctree(train, val);
 P=predict(T, chorale(1:20, :));
-finalResults=P;
+finalResults(1:20, 1)=P;
 
 train=chorale([1:20 41:100], :);
 val=validation([1:20 41:100], :);
 T=fitctree(train, val);
 P=predict(T, chorale(21:40, :));
-finalResults(21:40, :)=P;
+finalResults(21:40, 1)=P;
 
 train=chorale([1:40 61:100], :);
 val=validation([1:40 61:100], :);
 T=fitctree(train, val);
 P=predict(T, chorale(41:60, :));
-finalResults(41:60, :)=P;
+finalResults(41:60, 1)=P;
 
 train=chorale([1:60 81:100], :);
 val=validation([1:60 81:100], :);
 T=fitctree(train, val);
 P=predict(T, chorale(61:80, :));
-finalResults(61:80, :)=P;
+finalResults(61:80, 1)=P;
 
 train=chorale(1:80, :);
 val=validation(1:80, :);
 T=fitctree(train, val);
 P=predict(T, chorale(81:100, :));
-finalResults(81:100, :)=P;
+finalResults(81:100, 1)=P;
 
-confusionmat(finalResults, validation)
+% confusionmat(finalResults, validation)
 clear k val train
 j=0;
 over=0;
 under=0;
 for i=1:100
-    if finalResults(i)-validation(i)==0
+    if finalResults(i,1)-validation(i)==0
         j=j+1;
-    elseif finalResults(i)-validation(i)>0
+    elseif finalResults(i,1)-validation(i)>0
         over=over+1;
     else
         under=under+1;
     end
 end
-view(T,'Mode', 'Graph')
-computer_model=[over;j;under]
+% view(T,'Mode', 'Graph')
+computer_model=[under j over]
 clear i j under over P data ans train val computer_model

### 2018-04-25 Paper.tex
Adding tree diagram to paper
Added results of computer model to paper

@@
 \usepackage{setspace}
+\usepackage{graphicx}
+\graphicspath{{/}}
 \linespread{1.25}
@@
 In the final model, each chorale was used in a decision tree to predict the final pitch.
-70 random chorales from the data set were used as the training set, with the remaining 30 chorales left for testing.
-This model was able to correctly predict between 62 and 75 of the 100 chorales.
+80 random chorales from the data set were used as the training set, with the remaining 20 chorales left for testing.
+Five total trees were created, with a different subset of 20 each time; these five trees therby predicted over the entire data set.
+This model was able to correctly predict 55 of the 100 chorales.
+Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major.
+In some cases, there would be only one solution for chorales which were in A and E major, usually MIDI note 69 (A4).
+Overall, the most prevalent node checked the starking key signature.
+After key signature, the most prevalent nodes checked the pitch of the first note.
+While these covered the majority of the nodes within each tree, there were a number of outlier nodes.
+For instance, one fairly common outlier was built on the pitch of the 16th or 17th note in the melody line; this note is likely the end of the first phrase of the melody.
+Other nodes were related to the overall speed of the chorale; one node checked the length of the 15th note, and another checked if the 34th note of the melody was on the 34th beat or later.
+There may be a pattern apparent in the data, where faster chorales are more likely to end on a particular note than slower chorales.\\
+\begin{figure}[h]
+  \includegraphics[width=13.5cm]{Tree5}
+  \centering
+  \caption{One of the five decision trees generated by the computer model, with the nodes relabeled for readability.
+This tree was trained on chorales 1 through 80, and tested on chorales 81 through 100.}
+\end{figure}
 
+Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy.
+The confusion matrix for the computer model had fewer rows and columns than the human model.
+The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings.
+While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.
 \section{Conclusion}
+Although the computer model performed at a similar level to the human model
 Regardless of the results of the model, the relationship between computers and music continues to provide a challenging roadblock on the path to human-like artificial intelligence.

### 2018-04-25 final.m

@@ -236,30 +236,35 @@ val=validation(21:100, :);
 T=fitctree(train, val);
 P=predict(T, chorale(1:20, :));
 finalResults(1:20, 1)=P;
+view(T,'Mode', 'Graph')
 
 train=chorale([1:20 41:100], :);
 val=validation([1:20 41:100], :);
 T=fitctree(train, val);
 P=predict(T, chorale(21:40, :));
 finalResults(21:40, 1)=P;
+% view(T,'Mode', 'Graph')
 
 train=chorale([1:40 61:100], :);
 val=validation([1:40 61:100], :);
 T=fitctree(train, val);
 P=predict(T, chorale(41:60, :));
 finalResults(41:60, 1)=P;
+% view(T,'Mode', 'Graph')
 
 train=chorale([1:60 81:100], :);
 val=validation([1:60 81:100], :);
 T=fitctree(train, val);
 P=predict(T, chorale(61:80, :));
 finalResults(61:80, 1)=P;
+% view(T,'Mode', 'Graph')
 
 train=chorale(1:80, :);
 val=validation(1:80, :);
 T=fitctree(train, val);
 P=predict(T, chorale(81:100, :));
 finalResults(81:100, 1)=P;
+% view(T,'Mode', 'Graph')
 
 % confusionmat(finalResults, validation)
 clear k val train
@@ -275,6 +280,5 @@ for i=1:100
         under=under+1;
     end
 end
-% view(T,'Mode', 'Graph')
 computer_model=[under j over]
 clear i j under over P data ans train val computer_model

### 2018-04-25 Paper.tex
Added note about KNN model

@@
 80 random chorales from the data set were used as the training set, with the remaining 20 chorales left for testing.
-Five total trees were created, with a different subset of 20 each time; these five trees therby predicted over the entire data set.
+Five total trees were created, with a different subset of 20 each time; these five trees therby predicted over the entire data set.\\
+
 This model was able to correctly predict 55 of the 100 chorales.
+21 chorales were overestimated, and 24 chorales were underestimated.
 Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major.
@@
 While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.
+
+A similar process was used with a $k$ nearest neighbors approach, but the results were significantly worse than the classification tree.
+With $k=1$, only 13 chorales were predicted correctly; increasing the value of $k$ was only able to raise the correct predictions to the mid-20s.
 \section{Conclusion}
@@
 The amount of interdisciplinary cooperation that will eventually be necessary is staggering; from computer scientists and mathematicians to create the models, to musicians for the actual art, to psychologists and philosophers to determine how a computer can best ``learn.''
-
+\newpage
 \begin{thebibliography}{10}

### 2018-04-25 Paper.tex
Do extra types of computer models

@@
 \documentclass[12pt]{article}
-\usepackage{}
 %%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Methods}
-Three separate models were created for this experiment: a naive baseline model, a human-made model with predefined rules, and a computer-generated model.
+Three separate models were created for this experiment: a naive baseline model, a human-made model with predefined rules, and a computer-generated model built on classification trees.
+In addition, three variations of the computer model were built.
+Two used the same structure as the original, but with smaller subsets of the data; the third variation used a k-nearest-neighbors tree.
 
@@
 Overall, this approach was able to correctly predict the note in 49 of 100 chorales.
-Of the incorrect predictions, 36 chorales ended on a note below the predicted note and 15 ended on a note above the predicted note.\\
+Of the incorrect cases, 36 chorales predicted a note below the actual note and 15 predicted a note above the actual note.
+This model had a Mean Squared Error (MSE) of 26.94.\\
+
+\begin{figure}[h]
+  \includegraphics[width=13.5cm]{Naive}
+  \centering
+  \caption{A confusion matrix of the naive model. The green region represents overestimates, the blue region represents underestimates.}
+\end{figure}
 
@@
 This model predicted the endings of a total of 56 chorales correctly.
-While the number of overestimations was reduced by around 60\%, there was a 50\% rise in the number of underestimations.
-The minimal improvement of the model would suggest that the patterns created for this model were either incorrect or insufficiently described the musical patterns.
-In any case, there is much room for improvement for this model.
+Despite only a minor increase in correct pitches, the MSE decreased dramatically to 4.32.
+While the number of underestimations was reduced by around 60\%, there was a 50\% rise in the number of overestimations.
+This switch may be caused by a number of possibilities.
+Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model.
 
@@
 This model was able to correctly predict 55 of the 100 chorales.
-21 chorales were overestimated, and 24 chorales were underestimated.
+21 chorales were overestimated, and 24 chorales were underestimated, with a MSE of 15.93.
 Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major.
@@
 There may be a pattern apparent in the data, where faster chorales are more likely to end on a particular note than slower chorales.\\
+
+Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy.
+The confusion matrix for the computer model had fewer rows and columns than the human model.
+The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings.
+While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\
+
 \begin{figure}[h]
@@
 
-Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy.
-The confusion matrix for the computer model had fewer rows and columns than the human model.
-The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings.
-While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.
+A similar process was used with a $k$ nearest neighbors approach, but the results were significantly worse than the classification tree.
+With $k=1$, only 13 chorales were predicted correctly; increasing the value of $k$ was only able to raise the correct predictions to the mid-20s.\\
 
-A similar process was used with a $k$ nearest neighbors approach, but the results were significantly worse than the classification tree.
-With $k=1$, only 13 chorales were predicted correctly; increasing the value of $k$ was only able to raise the correct predictions to the mid-20s.
+In an attempt to prune satistically insignificant data, the tree model was run again with the starting time, length, and time signature removed.
+This variation predicted 54 chorale endings correctly, only one less than the original computer model, and had a MSE of 14.94.
+However, the split between underestimates and overestimates was worse; 30 chorales were underestimated and 16 were overestimated.
+Another variation was created with the same fields removed as above, but with an extra caveat that only notes under a fermata were used to predict the endings.
+This variation was still able to predict 50 chorales correctly, with a similar spread of over- and underestimates as the original computer model: 28 under, 22 over, MSE of 17.36.
+Clearly, the most useful predictors are found at the ends of musical phrases, but there are still insights to be gained from the other notes.
 \section{Conclusion}
-Although the computer model performed at a similar level to the human model
+Although the computer model performed at a similar level to the human model in total correct, the human model had greater precision on its incorrect guesses than any variation of the computer model.
+The classification trees that were generated had less to do with actual patterns of notes in the music, and had more in common with the naive model than the human model.
+Ideally, a computer that truly thought like a human would, in this scenario, look at more than just key signature.\\
+
+Future work in this topic could be done to study the remaining Bach chorales not included in the UCI data set, as well as studying similar works from Bach's contemporaries.
+This model could be used to measure how similar the works of two different composers are.
+Similar models could be constructed to look at music of a different style from classical choral compositions.
+In particular, the 12-bar blues provides a chord structure even more rigid than that of the common practice period, but comes with the improvisation and non-chord tones of jazz \cite{music_dict}.
+On the other hand, the pentatonic scale - most commonly associated with folk music from Scotland and China - has fewer possible choices for the final note but more ambiguous chord structure \cite{music_dict}.\\
+
 Regardless of the results of the model, the relationship between computers and music continues to provide a challenging roadblock on the path to human-like artificial intelligence.
@@
 \bibitem{computer_composer}
-  De Felice, Clelia, et al. "Splicing music composition." Information Sciences, vol. 385 - 386, 2017, pp. 196-212. OhioLINK Electronic Journal Center, doi:10.1016/J.INS.2017.01.004.
+  De Felice, Clelia, et al. ``Splicing music composition.'' Information Sciences, vol. 385 - 386, 2017, pp. 196-212. OhioLINK Electronic Journal Center, doi:10.1016/J.INS.2017.01.004.
 
@@
 
+\bibitem{music_dict}
+  Kennedy, Michael, and Joyce Bourne Kennedy. ``The Concise Oxford Dictionary of Music.'' Oxford University Press, 5th edition, 2007.
+
 \bibitem{computers}
-  Miranda, Eduardo R. Composing music with Computers. New York, NY: Routledge, 2014. Safari Books Online, http://0-proquest.safaribooksonline.com.dewey2.library.denison.edu\\/book/audio/9780240515670.
+  Miranda, Eduardo R.
+``Composing music with Computers.'' New York, NY: Routledge, 2014.
+Safari Books Online, http://0-proquest.safaribooksonline.com.dewey2.library.denison.edu\\/book/audio/9780240515670.
 \end{thebibliography}

### 2018-04-25 final.m

@@ -195,6 +195,9 @@ confusionmat(z, results(:,1))
 confusionmat(z, results(:,2))
 % confusionmat(N_results(:,1), N_results(:,2))
 % confusionmat(tmp(:,1), N_results(:,2))
+mse_Naive=sum((N_results(:,1)-N_results(:,2)).^2)/length(N_results)
+mse_Human=sum((tmp(:,1)-N_results(:,2)).^2)/length(tmp)
+
 clear n ans i j k
 
 %% Build computer model: For each note in the chorale, calculate the probability of that note being the ending pitch
@@ -236,7 +239,7 @@ val=validation(21:100, :);
 T=fitctree(train, val);
 P=predict(T, chorale(1:20, :));
 finalResults(1:20, 1)=P;
-view(T,'Mode', 'Graph')
+% view(T,'Mode', 'Graph')
 
 train=chorale([1:20 41:100], :);
 val=validation([1:20 41:100], :);
@@ -281,4 +284,172 @@ for i=1:100
     end
 end
 computer_model=[under j over]
-clear i j under over P data ans train val computer_model
+mse_Computer = sum((finalResults-validation).^2)/length(validation)
+
+clear i j under over P ans train val computer_model
+
+%% Build computer model variant: No starting time, length, or time signature
+clear chorale finalResults N_results results S_results T z tmp
+for j=1:100
+    k=2;
+    i=1;
+    while k<=length(data)
+        if data(j,k) == "st"
+            k=k+1;
+            k=k+2;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            k=k+2;
+            chorale(j,i) = data(j,k);
+            k=k+2;
+            i=i+1;
+            k=k+2;
+            chorale(j,i) = data(j,k);
+            i=i+1;
+        elseif j==1
+            chorale=str2double(chorale);
+            break;
+        else
+            break;
+        end
+        k=k+1;
+    end
+end
+clear i j k
+train=chorale(21:100, :);
+val=validation(21:100, :);
+T=fitctree(train, val);
+P=predict(T, chorale(1:20, :));
+finalResults(1:20, 1)=P;
+% view(T,'Mode', 'Graph')
+
+train=chorale([1:20 41:100], :);
+val=validation([1:20 41:100], :);
+T=fitctree(train, val);
+P=predict(T, chorale(21:40, :));
[…]

### 2018-04-26 Paper.tex
Finish first draft of essay

@@
 Finally, a Boolean flag is set to true if the note is underneath a fermata.
+Fermatas generally indicate the end of a musical phrase or chord progression, and instruct the performer to hold the note for as long as deemed necessary.
 
@@
   \centering
-  \caption{A confusion matrix of the naive model. The green region represents overestimates, the blue region represents underestimates.}
+  \caption{A confusion matrix of the naive model.
+The indices relate to the MIDI values of the predicted/actual notes.
+The green region represents overestimates, the blue region represents underestimates.}
 \end{figure}
@@
 
+\begin{figure}[h]
+  \includegraphics[width=13.5cm]{Human}
+  \centering
+  \caption{A confusion matrix of the human model. As in Figure 1, indices are MIDI pitches, the green region is overestimates, and the blue region is underestimates.}
+\end{figure}
+
 This model is intended to mimic how a passive listener might try to predict the ending.
@@
 Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model.
+There are likely some patterns which are not considered by the algorithm, where the two penultimate notes have two possible endings.
+In these cases, only the more prevalent one was incorporated into the algorithm.
+Finally, any chorale which uses the last two notes as part of a key change will not be accurately predicted, since this model uses the relative position within the current key.
 
@@
 80 random chorales from the data set were used as the training set, with the remaining 20 chorales left for testing.
-Five total trees were created, with a different subset of 20 each time; these five trees therby predicted over the entire data set.\\
-
-This model was able to correctly predict 55 of the 100 chorales.
-21 chorales were overestimated, and 24 chorales were underestimated, with a MSE of 15.93.
-Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major.
-In some cases, there would be only one solution for chorales which were in A and E major, usually MIDI note 69 (A4).
-Overall, the most prevalent node checked the starking key signature.
-After key signature, the most prevalent nodes checked the pitch of the first note.
-While these covered the majority of the nodes within each tree, there were a number of outlier nodes.
-For instance, one fairly common outlier was built on the pitch of the 16th or 17th note in the melody line; this note is likely the end of the first phrase of the melody.
-Other nodes were related to the overall speed of the chorale; one node checked the length of the 15th note, and another checked if the 34th note of the melody was on the 34th beat or later.
-There may be a pattern apparent in the data, where faster chorales are more likely to end on a particular note than slower chorales.\\
-
-Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy.
-The confusion matrix for the computer model had fewer rows and columns than the human model.
-The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings.
-While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\
+Five total trees were created, with a different subset of 20 each time; these five trees thereby predicted over the entire data set.\\
 
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=7cm]{Tree1}
+  \includegraphics[width=7cm]{Tree2}\\
+  \centering
+  \includegraphics[width=7cm]{Tree3}
+  \includegraphics[width=7cm]{Tree4}
+  \caption{The four other trees generated by the computer model. }
+\end{figure}
+
+This model was able to correctly predict 55 of the 100 chorales.
+21 chorales were overestimated, and 24 chorales were underestimated, with a MSE of 15.93.
+Looking at the various decision trees generated by the model, most trees contained the same node: checking whether the starting key was either A or E major.
+In some cases, there would be only one solution for chorales which were in A and E major, usually MIDI note 69 (A4).
+Overall, the most prevalent node checked the starting key signature.
+After key signature, the most prevalent nodes checked the pitch of the first note.
+While these covered the majority of the nodes within each tree, there were a number of outlier nodes.
+For instance, one fairly common outlier was built on the pitch of the 16th or 17th note in the melody line; this note is likely the end of the first phrase of the melody.
+Other nodes were related to the overall speed of the chorale; one node checked the length of the 15th note, and another checked if the 34th note of the melody was on the 34th beat or later.
+There may be a pattern apparent in the data, where faster chorales are more likely to end on a particular note than slower chorales.\\
+
+Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy.
+The confusion matrix for the computer model has one less row and column than both the naive model and 3 fewer rows and columns than the human model.
+The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings.
+While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\
+
+\begin{figure}[h]
+  \includegraphics[width=13.5cm]{confusionMat}
+  \centering
+  \caption{The confusion matrix generated by the computer model.
+As in Figure 1, indices are MIDI pitches, the green region is overestimates, and the blue region is underestimates.
+In comparison with the matrix in Figure 1, this matrix has one fewer row and column.}
+\end{figure}
+
 A similar process was used with a $k$ nearest neighbors approach, but the results were significantly worse than the classification tree.
@@
 
-In an attempt to prune satistically insignificant data, the tree model was run again with the starting time, length, and time signature removed.
+In an attempt to prune statistically insignificant data, the tree model was run again with the starting time, length, and time signature removed.
 This variation predicted 54 chorale endings correctly, only one less than the original computer model, and had a MSE of 14.94.
@@
 The classification trees that were generated had less to do with actual patterns of notes in the music, and had more in common with the naive model than the human model.
-Ideally, a computer that truly thought like a human would, in this scenario, look at more than just key signature.\\
+Ideally, a computer that truly thought like a human would, in this scenario, look at more than just key signature and the first pitch.\\
 

### 2018-05-08 Paper.tex
Add index to confusion matricies

@@
 
-Johann Sebastian Bach, as a product of the Baroque period, followed a set of composition guidelines which gave the Common Practice Period its name.
+Johann Sebastian Bach, as a product of the Baroque period (1600-1750), followed a set of composition guidelines which gave the Common Practice Period its name.
 These guidelines dealt with methods of resolving dissonance, writing melodies, and more.
@@
 There have been a number of attempts to create algorithmic composition software, a program which can compose music with little human interaction or involvement; one of the first examples dating to the 1950's when Lejaren Hiller and Leonard Isaacson produced {\it The Illiac Suite for String Quartet} \cite{computers}.
-There are two varieties of algorithmic composition: top-down and bottom-up; an implementation of the bottom-up approach is covered in detail later.\\
+There are two varieties of algorithmic composition: top-down and bottom-up.\\
 
@@
 \subsection{A Bottom-Up Approach: Splicing Systems}
-In their attempt to create an bottom-up algorithmic composition model, De Felice et.
+In their attempt to create a bottom-up algorithmic composition model, De Felice et.
 al.
@@
 \section{Data}
-The data set used to test the models is a collection of one hundred Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}.
+The data set used to test the models is a collection of 100 Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}.
 Each chorale contains only the melody line from each chorale.
@@
 The starting time of the note is measured in 16th notes from the start of the piece.
-The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C\#4 given a value of 61, etc.
+The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C$^{\#}$4 (C sharp) given a value of 61, etc.
 The pitches are restricted to values between 60 and 75, by nature of the notes present in the chorales.
@@
 In addition, three variations of the computer model were built.
-Two used the same structure as the original, but with smaller subsets of the data; the third variation used a k-nearest-neighbors tree.
+Two used the same structure of classification trees as the main model, but with smaller subsets of the data; the third variation used a k-nearest-neighbors tree.
 
@@
 With these values calculated, the model uses a defined set of patterns to predict the melody endings.
-For the sake of simplicity, only notes in the major triad were used in this model's predictions.\\
+For the sake of simplicity, only notes in the major triad (tonic, major third, perfect fifth) were used in this model's predictions.\\
 
@@
 Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model.
-There are likely some patterns which are not considered by the algorithm, where the two penultimate notes have two possible endings.
+There are likely some patterns which are not considered by the algorithm, cases where the same pattern of two notes has two possible endings.
 In these cases, only the more prevalent one was incorporated into the algorithm.
@@
 The confusion matrix for the computer model has one less row and column than both the naive model and 3 fewer rows and columns than the human model.
-The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings.
+The pitches for the notes were given in a range between 60 (C4) and 75 (E$^b$5) in the MIDI scale, but this range applies to the entire melody line, not the endings.
 While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\
